---
title: "Lab 6 - Model selection"
author: "Ilaria Prosdocimi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
---
  
We will analyze the dataset `prostate`: 
  
```{r}
urlLocation <- "https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.data"
prostate <- read.table(urlLocation, header=TRUE)[,1:9]
## explanation of the data 
## https://web.stanford.edu/~hastie/ElemStatLearn/datasets/prostate.info.txt
## save the data locally as 
## write.csv(x = prostate, file = "prostate.csv", row.names = FALSE, quote = FALSE)
## so you can re-read as 
## prostate <- read.csv("prostate.csv", header = TRUE)
```

The variables in the dataset are

* lcavol: log cancer volume
* lweight: log prostate weight
* age: age of patient
* lbph: log of the amount of benign prostatic hyperplasia
* svi: seminal vesicle invasion (binary variable)
* lcp: log of capsular penetration
* gleason: Gleason score
* pgg45: percent of Gleason scores 4 or 5
* lpsa: log PSA score


We aim to select a model that to explain how `lpsa` is affected by other variables: this might be useful both for prediction purposes (i.e. predict the `lpsa` for a new patient) and for inference purposes (i.e. understanding which factors influence the `lpsa` values).  

The relationship of `lpsa` against all the other variables in the dataset can be seen in the plot below: 

```{r plotMatrix}
plot(prostate)
```


```{r}
fit_minimum <- lm(lpsa ~ 1, data = prostate)
fit_maximum <- lm(lpsa ~ ., data = prostate)
```

```{r}
AIC(fit_minimum)
AIC(fit_maximum)
```

```{r}
for(j in 1:8){
  print(c(names(prostate)[j],AIC(lm(prostate$lpsa~prostate[,j]))))
}
fit_for1 <- lm(lpsa ~ lcavol, data = prostate)
## end of step 1
```


```{r}
AIC(fit_for1)
for(j in 2:8){
  print(c(names(prostate)[j],
          AIC(lm(prostate$lpsa~prostate$lcavol + prostate[,j]))))
}
fit_for2 <- lm(lpsa ~ lcavol+lweight, data = prostate)
## end of step 2
```


```{r}
AIC(fit_for2)
for(j in 3:8){
  print(c(names(prostate)[j],
          AIC(lm(prostate$lpsa~prostate$lcavol +
                               prostate$lweight + 
                               prostate[,j]))))
}
fit_for3 <- lm(lpsa ~ lcavol+lweight+svi, data = prostate)
## end of step 3
```



```{r}
AIC(fit_for3)
for(j in c(3,4,6,7,8)){
  print(c(names(prostate)[j],
          AIC(lm(prostate$lpsa~prostate$lcavol +
                               prostate$lweight +
                               prostate$svi + 
                               prostate[,j]))))
}
fit_for4 <- lm(lpsa ~ lcavol+lweight+svi+lbph, data = prostate)
## end of step 4
```



```{r}
AIC(fit_for4)
for(j in c(3,6,7,8)){
  print(c(names(prostate)[j],
          AIC(lm(prostate$lpsa~prostate$lcavol +
                               prostate$lweight +
                               prostate$svi +
                               prostate$lbph +
                               prostate[,j]))))
}
fit_for5 <- lm(lpsa ~ lcavol+lweight+svi+lbph+age, data = prostate)
## end of step 5
```



```{r}
AIC(fit_for5)
for(j in c(6,7,8)){
  print(c(names(prostate)[j],
          AIC(lm(prostate$lpsa~prostate$lcavol +
                               prostate$lweight +
                               prostate$svi +
                               prostate$lbph +
                               prostate$age +
                               prostate[,j]))))
}
## end of step 6
```


```{r}
step(object = fit_minimum, 
     scope = list(upper = fit_maximum, 
                  lower = fit_minimum), 
     direction = "forward", 
     # usiamo AIC 
     k = 2)
```



```{r}
aic_bsel <- 
  step(object = fit_maximum, 
     scope = list(upper = fit_maximum, 
                  lower = fit_minimum), 
     direction = "backward", 
     # usiamo AIC 
     k = 2)
```

Cosa succede se usiamo il BIC come criterio di scelta? 
Il BIC si differenzia dall'AIC per l'uso di log(n) invece che 2 per penalizzare la complessità del modello. 



```{r}
step(object = fit_minimum, 
     scope = list(upper = fit_maximum, 
                  lower = fit_minimum), 
     direction = "forward", 
     # usiamo BIC 
     k = log(nrow(prostate)))
```



```{r}
bic_bsel <- 
  step(object = fit_maximum, 
     scope = list(upper = fit_maximum, 
                  lower = fit_minimum), 
     direction = "backward", 
     # usiamo BIC 
     k = log(nrow(prostate)), 
     trace = 0)
```




```{r}
step(object = lm(lpsa ~ lcavol, data = prostate), 
     scope = list(upper = fit_maximum, 
                  lower = fit_minimum), 
     direction = "both", 
     # usiamo BIC 
     k = log(nrow(prostate)))
```


## Exhaustive search 


```{r}
# install.packages("leaps")
library(leaps)
```

```{r}
rg_1 <- regsubsets(lpsa~., data = prostate)
summary(rg_1)
summary(rg_1)$rsq
summary(rg_1)$adjr2
plot(seq(1, 8), summary(rg_1)$adjr2, type = "l")
summary(rg_1)$which
```


## Validazione incrociata (leave one out cross validation) 


```{r}
loo_cv_vals <- NULL  # e_[i]
for(i in 1:nrow(prostate)){
  refit <- lm(formula = lpsa ~ lcavol + lweight + svi, 
              data = prostate[-i,])
  loo_cv_vals <- c(loo_cv_vals, 
                   (prostate[i,"lpsa"] - 
                     predict(refit, newdata = prostate[i,])))
}
length(loo_cv_vals)
sum(loo_cv_vals^2)/nrow(prostate)

## now for the model selected by AIC
loo_cv_vals <- NULL  # e_[i]
for(i in 1:nrow(prostate)){
  refit <- lm(formula = lpsa ~ lcavol + lweight + age + lbph + svi, 
              data = prostate[-i,])
  loo_cv_vals <- c(loo_cv_vals, 
                   (prostate[i,"lpsa"] - 
                     predict(refit, newdata = prostate[i,])))
}
length(loo_cv_vals)
sum(loo_cv_vals^2)/nrow(prostate)
```


```{r}
X <- model.matrix(bic_bsel)
Hmat <- X %*% solve(t(X) %*% X) %*% t(X)
yhat <- Hmat %*% prostate$lpsa
sum((residuals(bic_bsel)/(1-diag(Hmat)))^2)/nrow(prostate)
```

```{r}
calc_loo_rmse <- function(model){
  sqrt(mean((residuals(model)/(1-hatvalues(model)))^2))
}
calc_loo_rmse(bic_bsel)
```

# Verificare le assunzioni del modello 

Omoschdasticità 

```{r}
plot(fitted(bic_bsel), resid(bic_bsel), pch = 16)
```

Va bene se... non c'è relazione evidente e la varianza è costante. Sembra OK. 

Normalità 


```{r}
qqnorm(resid(bic_bsel)); qqline(resid(bic_bsel))
car::qqPlot(bic_bsel)
```

